Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-11 probabilisticProgrammingMartingales.tex
Fix some examples.

@@

Programs $M_1$ and $M_3$ are AST.
To construct antitone ranking functions, we appeal to a partial ranking function theorem for antitone ranking functions.
\iffalse
\lo{NOTE. Suppose a function $f : E \to \Real$ from a convex subset $E \subseteq \Real^n$ has continuous second-order partial derivatives. Then it is convex iff its Hessian is positive semidefinite.}
\fi
\iffalse
For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
\fi

For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
\[
\left\{
\begin{array}{rll}
z_0 &=& 0\\
n \geq 1, \quad z_n &>& \dfrac{1}{2} \, \big( z_{n-1} + z_{n+1}\big).
\end{array}
\right.
\]
Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
Then
\begin{equation}
\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
\label{eqn:epsilon g M1}
\end{equation}
(Since $g_1$ is bijective, with inverse $g_1^{-1}(y) = 1 - \log (2-y)$, we can obtain $\epsilon_1(y)$ from \Cref{eqn:epsilon g M1}.) 
%We have $1/2 \cdot \big(g_1(x-1) + g_1(x+1)\big) = 2 - \frac{5}{2^{x+1}}$.

Using shorthand 
$\Theta_1 = \tY \, f \, n \, . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define a partial antitone ranking function $f_1 : \mathit{Rch}(M_1) \to \nnReal$ with antitone function $\epsilon_1$ defined as follows:
\begin{align*}
{\Theta_1} \, n 
&\mapsto 
g_1(n)
\\
\tif{n = 0}{0}{\Theta_1 \, (n - 1) \oplus_{1/2} \Theta_1 \, (n + 1)}
&\mapsto
\dfrac{1}{2} \, \big(g_1(n-1) + g_1(n+1)\big)
\\
0 &\mapsto 0
\end{align*}

For $M_3$, we use the function $g_3(x) := 3 - \frac{1}{3^{x-1}}$ and the antitone function $\epsilon_3(y) := \frac{84-28y}{27}$. 
Then
\[
\epsilon_3(g_3(x)) = g_3(x) - \dfrac{1}{3}\big(2 \cdot g_3(x-1) + \cdot g_3(x+2) \big) = 
\frac{28}{3^{x+2}}.
\]
Following $M_1$ above, we can define a partial antitone ranking function $f_3 : \mathit{Rch}(M_3) \to \nnReal$ with $\epsilon_3$ as the antitone function.

>>>>

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values). Consider, for example, a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T = $ the least $n$ such that $2^{-n} < X$, $Y_k = 4-2^{-k}$ if $k < T$, or $0$ otherwise, and $\epsilon(x) = \frac{4-x}{4}$. In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. It is an antitone strict supermartingale except that $\epsilon$ goes to $0$ at $4$, which is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $T = \infty$ with probability $\frac 1 2$.

@@

Programs $M_1$ and $M_3$ are AST.
To construct antitone ranking functions, we appeal to a partial ranking function theorem for antitone ranking functions.

For $M_1$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_1 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_1 : \mathit{Rch}(M_1) \rightharpoonup \nnReal$ with antitone function $\epsilon_1$ defined as follows (for $n \in \mathbb N$):
\begin{align*}
{\Theta_1} \, n 
&\mapsto 
g_1(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_1 \, n$ reduces in several steps to either $\Theta_1 \, (n-1)$ or $\Theta_1 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
\begin{align*}
  & g_1(n) - \frac{g_1(n-1) + g_1(n+1)} 2 \\
  = & \log(\frac{n+1}{\sqrt{n(n+2)}}) \\
  = & \frac 1 2 \log(\frac{(n+1)^2}{n(n+2)}) \\
  = & \frac 1 2 \log(1 + \frac 1 {n(n+2)}) \\
  > & \frac 1 {4n(n+2)} \\
  > & \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
and $\Theta_1 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_1$ of $1$, therefore by setting $\epsilon_1(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_1$ must decrease is met. It is also defined at $M_1 = \Theta_1 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_1$ is AST.
